The difference operation between templates of binary cellular automata

### 2015-11-19 article/article.tex
Revision by Verardo

\begin{abstract}
The templates consist in a notation that has recently been introduced to represents families of cellular automata that share some property. This paper introduces the operation of difference between templates. This operation is applicable to representation of binary cellular automata families that share some properties but not others. This papers also introduce a process called \textit{exception templates} that is required to this operation. Experimental results and examples of both techniques defined above are illustrated with examples in the space of elementary cellular automata.
\keywords{cellular automata, rule space, templates, difference between templates, exception templates.}
\end{abstract}
%
\section{Introduction}
\label{sec:introducao}
Cellular automata (CAs) are dynamical systems typically discrete in time, space and states.
CAs can produce behavior of great complexity based on simple rules of local action \cite{wolfram2002}. The study of classical problems of cellular automata, as the problem of parity \cite{Betel2013} and the density problem \cite{deOliveira2014density}, can help in the understanding of how this complex behavior emerges.

The most basic approach to find a suitable solution to classic problems consist in test each of rules of a particular family of CAs in order to see if any of these rules are able to solve the problem. However, this approach shows to be inefficient or impractical for a large number of families in CAs, which is the most usual panorama in literature.

Evolutionary computing is constantly used to deal with larger families. This type of algorithm have proven itself very effective in density classification problem \cite{wolz2008very}.

Another strategy to find suitable rules is restricting the search space by rules that have a given property. To represent a subspace with a particular property without the need of enumerate all the rules contained therein can use \textit{templates} idea, as proposed in \cite{deOliveira2014,deOliveira2014b}.

There are a lot of operations that can be applied to templates such as expansion and intersection. In this paper we introduce the operation named \textit{difference between templates} and the operation that generates the \textit{exception templates}. Moreover, application of these operations are presented in this paper with practical examples.

@@

\section{Cellular Automata}
\label{sec:automatos_celulares}
Cellular Automata are simple mathematical idealizations of natural systems \cite{wolfram1994cellular}. They consists of a lattice of discrete fields usually identical, called cells, each of which can take on a finite set of generally integers. The values (or states) of the cells evolve in discrete time steps usually according to deterministic rules that specify the state of each cell according to the states of their neighboring cells \cite{wolfram1994cellular}.% "Cellular Automata are simple mathematical idealizations of natural systems" está identico ao autor.

Conventionally, it is assumed that a cellular automaton cells have $k$ states which are represented by integer values in the range of $[0, k-1]$ The state of a cell is changed by the local function of the automaton (a rule), formed by the state transitions set of a cell, based on its current state and in the states of adjacent cells. To update the values of a cell by local functions, usually a radius of action $r$ is defined and $r$ is the number of adjacent cells to be analyzed in each direction by local functions.

A family of cellular automata, or space of cellular automata, is defined by the radius $r$ and the number of states $k$. One-dimensional cellular automata with $r=1$ e $k=2$ are called the family of elementary cellular automata.

@@

The number of rules in a space is provided by $k^{k^{2r+1}}$, making it easy to realize that any change in the variable $k$ and $r$ generate families with a large amount of rules. An approach to handle this problem is to use static properties. The static properties consists of properties obtained directly from the set of CA state transitions. The use of static properties can highly restrict the initial search space. With templates can be represented a set of rules with certain static property.

However, in order to explain the functioning of templates and its operations it is interesting understand the properties of number conservation, internal symmetry and color blind, because these properties will be used as an example later.

\subsection{Number Conservation}
Number conservation is a static property that determines that the sum from the states of a particular cellular automata should not change during the evolution of space-time, regardless of the initial configuration.

According to Boccara and Fukś \cite{boccara2002}, a CA is conservative when each of its local function $f$, applied in every neighborhood $(\alpha_0,\alpha_1, \dots, \alpha_{n-1})$ satisfy the conditions described in Eq. \ref{eq:conservativeCA}.
\begin{equation}
\begin{split}
f(\alpha_0,\alpha_1, \dots,\alpha_{n-1}) = \alpha_0 + (\sum_{i=0}^{n-2}f(0_0,0_1, \dots,0_i,\alpha_1,\alpha_2, \dots,\alpha_{n-1}) \\- f(0_0,0_1, \dots,0_i,\alpha_0,\alpha_1, \dots,\alpha_{n-i-1}))
\label{eq:conservativeCA}
\end{split}
\end{equation}

>>>>

\begin{abstract}
A cellular automata template consist in a notation that has recently been introduced to represent families of cellular automata that share a given property. This paper introduces the operation of difference between templates, used to find cellular automata that have a given property but lack another one. This paper also introduce a process called \textit{exception templates} that is required to this operation. Experimental results of both techniques are illustrated with examples in the space of elementary cellular automata.
\keywords{cellular automata, rule space, templates, difference between templates, exception templates.}
\end{abstract}
%
\section{Introduction}
\label{sec:introducao}
Cellular automata (CAs) are dynamical systems typically discrete in time, space and states.
CAs can produce behavior of great complexity based on simple rules of local action \cite{wolfram2002}. The study of classical problems of cellular automata, like the parity problem \cite{Betel2013} and the density classification problem \cite{deOliveira2014density} can help to understand how this complex behavior emerges.

The most basic approach to find a suitable solution to classic problems consists in testing each of the rules of a particular family of CAs in order to see if any of these rules are able to solve the problem. However, this approach shows to be inefficient or impractical for larger families of CAs, which is the most usual panorama in literature.

Evolutionary computing is constantly used to deal with larger families. This type of algorithm has proven very effective to find solutions for the density classification problem \cite{wolz2008very}.

Another strategy to find suitable rules is restricting the search space by rules that have a given property. To represent a subspace with a particular property without the need of enumerate all of the space rules, one can use \textit{templates}, as proposed in \cite{deOliveira2014,deOliveira2014b}.

There are two operations that can be applied to templates: \textit{expansion} and \textit{intersection} \cite{deOliveira2014,deOliveira2014b}. In this paper we introduce a new operation, named \textit{difference between templates} and the operation that generates the \textit{exception templates}. Moreover, application of these operations are presented in this paper with practical examples.
% Atualmente, não temos "a lot of operations", só temos a expansão e a interseção. Pode dizer isso e citar os artigos onde elas estão explicadas. Tente sempre evitar termos meio vagos como "a lot of" e etc.
% Po, "difference between templates" é um nome meio grande...hahaha será que não tem outro melhor?

@@

\section{Cellular Automata}
\label{sec:automatos_celulares}
Cellular Automata are simple mathematical idealizations of natural systems \cite{wolfram1994cellular}. They consist of a lattice of discrete cells, each of which can be on one of a finite set of states in a given time step. The values (or states) of the cells evolve in discrete time steps usually according to deterministic rules that specify the state of each cell according to the states of their neighboring cells \cite{wolfram1994cellular}.% "Cellular Automata are simple mathematical idealizations of natural systems" está identico ao autor. "which can be on one of a"?

Conventionally, it is assumed that a cellular automaton cells have $k$ states which are represented by integer values in the range of $[0, k-1]$ The state of a cell is changed by the local function of the automaton (a rule), formed by the state transitions set of a cell, based on its current state and in the states of adjacent cells. To update the values of a cell by local functions, usually a radius $r$ is defined as the number of adjacent cells to be analyzed in each direction by local functions.

A family (space) of cellular automata, is defined by the radius $r$ and the number of states $k$. One-dimensional cellular automata with $r=1$ e $k=2$ are called the family of elementary cellular automata.

@@

The number of rules in a space is provided by $k^{k^{2r+1}}$, making it easy to realize that any increase in the values of $k$ and $r$ generates families with a large amount of rules. An approach to handle this problem is to use static properties as a hint on how the CA will behave during its evolution. A static property is obtained directly from the CA's rule table. The use of static properties can highly restrict the initial search space. With templates can be represented a set of rules with certain static property.

In order to explain how templates work, it is important to understand the properties of number conservation, internal symmetry and color blindness, that will be used in later examples.

\subsection{Number Conservation}
Number conservation is a static property that determines that the sum of the states of a particular cellular automata should not change during the CA's evolution, regardless of the initial configuration.

According to Boccara and Fukś \cite{boccara2002}, a CA is conservative when each of its local function $f$, applied in every neighborhood $(\alpha_0,\alpha_1, \dots, \alpha_{n-1})$ satisfies the conditions described in Eq. \ref{eq:conservativeCA}.
\begin{equation}
\begin{split}
f(\alpha_0,\alpha_1, \dots,\alpha_{n-1}) = \alpha_0 + (\sum_{i=0}^{n-2}f(0_0,0_1, \dots,0_i,\alpha_1,\alpha_2, \dots,\alpha_{n-1}) \\- f(0_0,0_1, \dots,0_i,\alpha_0,\alpha_1, \dots,\alpha_{n-i-1}))
\label{eq:conservativeCA}
\end{split}
\end{equation}
